LHC abort gap monitor electronics upgrade

### 2024-04-25 paper.tex
rephrased some chapters

\subsection{Current Solution}
The current solution uses DAB64x card deploying FPGA Altera Stratix EP1S40. The
platform has two FMC connectors. There are mounted two IBMS module. Those
modules integrate voltage on input using ASIC lhcb2002b~\cite{lhcb_asic}.
To get the maximal resolution of the ADC, the signal is amplified by HAMAMATSU
C5594 with static gain \SI{36}{dBV}, bandwith \SI{1.5}{GHz} and the low cutoff
frequency 50kHz.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{existing_system_processing.pdf}
    \caption{Current system block diagram}
    \label{fig:current_system_processing}
\end{figure}
The system schematic is in Fig.~\ref{fig:current_system_processing}. The OF is
the same as described in the system architecture chapter. The picture omits the
filters, because they are used in all cases. The MCP-PMT signal needs to be
amplified by at least 34dB to use the full scale of the 40Mhz ADC. The
Hamamatsu C559 is used~\cite{bsra_first_operation}. The signal belonging
to one bunch (\SI{25}{ns}) is integrated in the IBMS (blue triangle with a
capacitor in the feedback) sampled by ADC (violet rectangle in the figure). The
DAQ integrates 4 samples (light blue rectangle with $\sum$ symbol) to calculate
charge in \SI{100}{ns} interval. It produces 64 values, from which are created
two sets of 64 values with the standard deviation and the mean value is
calculated. Both sets are passed to the EMA block (green rectangle in
the figure). All data are saved in the block ram. This is is than shared with
the PC over wishbone to VME64 bridge. The PC runs server which provides the
FESA interface to the connected clients.

\subsection{Constraints and Limitations}
The current AGM returns artefacts in the measurements. The signal, which lays
outside of the gate is not reflecting behaviour of white noise. This is caused
by the lhcb2002b ASIC. The integrators gain is not linear, have different
offset and dead times. This means, that system needs to correct at each
integrator these effects, which creates uncertainties in the results. This
results in complicated calibration procedur.\\
The system uses old FPGA platform for which development sofware tools are not
compatible with newest operation systems. The amplifier is not variable, which
results in lower resolution for smaller magnitudes.

\section{System Upgrade}
To resolve the issue of the maintenance, the VFC-HD carrier is used with
additional RTM board providing driver for the gate and control of the built-in
test LED. The same carrier board with 500MSPS 14-bit ADC has been used in other
projects and so it provided the requirement for reducing the maintenance efford.
The new system removes IBMS cards with lhcb2002b integrators and moving the
whole integration process to the gateware.

\subsection{Low band Amplifier}
The new custom amplififer is designed to allow work with higher signal dynamic
range, because of its variability.  
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))
In the Fig. \ref{fig:amp_schematics} is shown amplifier's block scheme. All the
settings is controlled via UART interface. It provides changing the variable
current-to-voltage conversion factor from \SI{5.8}{mV/\micro A} -
\SI{1.1}{V/\micro A}. This corresponds to the Voltage gain from \SI{41}{dB} to
\SI{87}{dB}. The AGM ADC has a no zero offset, so the signal can be shifted in the
amplifier before post-processing.\\
The amplifier features also alarm signal generator. This serves to turn of the
gate, when the light impinging is to intense and it could damage the
multi-channel plate inside the PMT. The alarm signal is send to the gateware,
where it is handled. The alarm feature has a variable threshold set.   
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier}
    \label{fig:amp_schematics}
\end{figure}
The amplifier design has three switches. First switch can set pre-integrator,
to filter out spikes caused by dark counts of the PMT. Second switch can
activate either manual or gate driven integration of the acquisition window.  
Last switch activates low pass filtering with cutoff frequency of 40MHz.

\subsection{Acquisition system}
The DAQ had to be redesigned to be compatible with the new VFC-HD carrier. It
needed Rear transition module (RTM) to control the gate and built-in test LED.  
The Fig.~\ref{fig:bsra_daq} is depicted DAQ functionality principle. 
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_daq.pdf}
    \caption{Ackquisition system schematics}
    \label{fig:bsra_daq}
\end{figure}
The new DAQ gets asynchronous data samples from ADC 500MSPS 14-bit (violet
rectangle) in the block of 4 frames every \SI{8}{\nano s} belonging to one
channel. Each frame has 4 x \SI{16}{bit} values. To calculate the integrals of
\SI{100}{\nano\second}, the system tags the data. The tagging block resamples
LHC timing by 500MSPS and creates bunch and turn clock vector masks. Those
store the information about both clocks phases. The 14-bit data from ADC are
left aligned to 16-bit values. The tagger sets their two LSBs to one, when the
ADC sample falls into the edge range. The system stores data from region of
interest, into the FIFO using the synchronization block pulse. Stored data are
then processed as a stream of values passed to the integrator. This starts
summing up values, when it hits the tag in the data indicating the start of the
bunch. It creates 64 values offseted from the pedestal memory (P[63:0]). This
subtracts accumulated charge at the gate opening and closing edge. The values
are then sent to the EMA and the result stored in the block ram available on
the VME64x bus. The format of the data provided to the PC with runnig FESA
server is the same with the current AGM~\cite{my_thesis}.

\subsection{Installation in LHC}
The instrument front and Acquistion system is located in LHC Point 4 near RF
system. The new installation is shown in Fig.~\ref{fig:installation}
highlighted with violet frame. On its right side is the currently used AGM
and on the left side is the BSRS. All the instruments share the same light
source with the BSRT. The instrument is pulled out from the shielding and the
red arrow ilustrates, where the light comes from and goes in to the MCP-PMT of
the new AGM.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{installation.jpg}
    \caption{New AGM installed in Point 4 next to BSRT}
    \label{fig:installation}
\end{figure}
On top of the reflective foil directing light towards the MCP-PMT
is a LED simulating the light source. This helps verifying the system
operation. The AGM amplifier is placed on top of the optical front-end to be
reachable in case of intervention. For reducing the light intensity, there are
pneumatically controlled optical filters in front of the MCP-PMT.

\section{Control Interface}
The gateware outputs data from four different block rams. The first stores raw
data from the ADC, that is used correctly set the initial delay of the
instrument's timing. It stores 1.5 turns and therefore it is published with
\SI{3.3}{Hz}. The next two memories store the actual snapshot data and averaged
data, but with finer granularity. They store 10 samples per \SI{10}{ms} window,
which gives the sample rate of \SI{100}{Hz}. The last memory is just bare
averaged data taken every \SI{10}{ms} defining the publishing rate. The
fetching handles the FESA server, which obtains interrupt every time the data
are ready. 
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{control_interface.jpg}
    \caption{New AGM expert graphical interface}
    \label{fig:control_interface}
\end{figure}
To simplify the setup tunining a new interface had to be created. It runs as
client of FESA server and displays the content of all the block rams. The main
control panel is depicted in Fig.~\ref{fig:control_interface}. It displays
three plots, raw data from the ADC (top plot), processed averaged data (down
plot on the left) and and evolution of the data in time (down plot on the
right). The menu on the left controls the amplifier and bottom menu controls
the acquisition delays and optical filters.

\section{First acquired plots}
First measurements were taken in November 2023 on the ion beam. The
optical system splits the SL in ratio 1:8, so the new system gets
significantly less light then current one. In other word, we expect
lower value deviation for system measuring light with higher intensity.  
\begin{figure}[!htb]
    \begin{center}
        \scalebox{0.54}{\input{plots/ions.pgf}}
        \caption{Features comparison of AGMs}
        \label{fig:comparison_chart}
    \end{center}
\end{figure}
The Fig. \ref{fig:comparison_chart} shows first acquisition of the new system
in comparison with the current one. The red plot is measured by the current
system and the blue chart is the new acquisition. In a range from 47th - 63th
bin, the gate is already close. There is just white noise, but the red plot
slowly decays and it continues below the zero line. This artefact is caused by
analog integrators and it is non-linear. The new system removes this effect.
The value deviation of the new acqusition from the current system represents
the yellow plot. The relative value deviation of the new system from the
current one is depicted with the yellow plot. In this measurement the maximum
deviation is 4\% in the area of the decay of the red plot. The the deviation of
2\% in the steep parts can be caused by the sample phase alignment of the two
systems.

\section{Summary}
The new AGM has been implemented on the CERN BI standard platform VFC-HD
carrier simplifying maintenance. The design removed the IBMS cads with ASIC
lhcb2002b and so resolved the issue with cumbersome calibration and artefacts
in the output plots.  It also The changes in the AGM design compared to the
current AGM are summarized in the following table Tab.
\ref{tab:system_comparison}.
\input{tables/system_comparison.tex} 
The system is installed in LHC the point 4 next to the current AGM collecting data
to be further analyzed. For the complet reliability test and proper function
verification, it needs to be collected more data, that will help evalate system
goodness. The systems shows so far, that with less light, it can have still
comparable outputs to the current AGM.

>>>>

\subsection{Current Solution}
The current setup employs a DAB64x card with an FPGA Altera Stratix EP1S40 for
data processing. The platform features two FMC connectors, each hosting an IBMS
module integrating voltage input via the lhcb2002b ASIC~\cite{lhcb_asic}. To
maximize ADC resolution, the signal undergoes amplification using a HAMAMATSU
C5594 amplifier with a static gain of \SI{36}{dBV}, a bandwidth of
\SI{1.5}{GHz}, and a low cutoff frequency of \SI{50}{Hz}.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{existing_system_processing.pdf}
    \caption{Current system block diagram}
    \label{fig:current_system_processing}
\end{figure}
The system schematic is depicted in Fig.\ref{fig:current_system_processing}.
The OF remains consistent with the description provided
in the system architecture chapter. Notably, filters are not represented in the
image as they are universally applied. For optimal utilization of the \SI{40}{MHz}
ADC's full scale, the MCP-PMT signal necessitates a minimum amplification of
\SI{34}{dB}, a task accomplished by the Hamamatsu C559
amplifier\cite{bsra_first_operation}.\\
Within the schematic, the signal corresponding to a single bunch (\SI{25}{ns})
is integrated within the IBMS module (denoted by a blue triangle with a
capacitor in the feedback loop) and subsequently sampled by the ADC
(represented by a violet rectangle). The DAQ system integrates four samples
(indicated by a light blue rectangle with a $\sum$ symbol) to compute the
charge over a \SI{100}{ns} interval. This process generates 64 values, from
which two sets of 64 values are derived, encompassing the standard deviation
and mean value calculations. Both sets are then forwarded to the Exponential
Moving Average (EMA) block (depicted by a green rectangle). All collected data
are stored within the block RAM and subsequently shared with a PC via a
Wishbone to VME64 bridge. The PC operates a server, facilitating the FESA
interface for connected clients.

\subsection{Constraints and Limitations}
The present AGM exhibits artifacts in the measurements. Signals outside the
gating range do not exhibit the characteristics of white noise, attributed to
the lhcb2002b ASIC. The integrators' gain lacks linearity, featuring distinct
offsets and dead times. Consequently, the system necessitates correction for
these effects at each integrator, leading to uncertainties in the outcomes and
necessitating intricate calibration procedures.\\
The system relies on an outdated FPGA platform, rendering its development
software incompatible with the latest operating systems. Additionally, the
fixed amplifier configuration contributes to reduced resolution for signals of
smaller magnitudes.

\section{System Upgrade}
To address maintenance concerns, the VFC-HD carrier is employed alongside an
additional RTM board, facilitating gate driver functionality and control of the
built-in test LED. Leveraging the same carrier board equipped with a 500MSPS
14-bit ADC from previous projects meets the requirement for reducing
maintenance efforts. The updated system eliminates IBMS cards featuring
lhcb2002b integrators, transitioning the entire integration process to the
gateware.

\subsection{Low band Amplifier}
The newly developed custom amplifier is designed to accommodate a wider signal
dynamic range by virtue of its variability.
% A = U^2/(i^2 * R^2) => A_dB = 20*log10(U/(i*R))
Figure \ref{fig:amp_schematics} illustrates the block diagram of the amplifier.
All configurations are managed through a UART interface, facilitating
adjustments to the variable current-to-voltage conversion factor ranging from
\SI{5.8}{mV/\micro A} to \SI{1.1}{V/\micro A}. This corresponds to a voltage
gain range of \SI{41}{dB} to \SI{87}{dB}. The AGM ADC does not exhibit a zero
offset, allowing for signal shifting within the amplifier prior to
post-processing.\\
Additionally, the amplifier incorporates an alarm signal generator designed to
deactivate the gate when the incident light intensity exceeds safe levels,
potentially risking damage to the multi-channel plate within the PMT. The alarm
signal is transmitted to the gateware for processing and management. Notably,
the alarm feature includes a variable threshold setting.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{amp_schematics.pdf}
    \caption{New AGM custom amplifier}
    \label{fig:amp_schematics}
\end{figure}
The amplifier design includes three switches. The first switch can configure
the pre-integrator to filter out spikes resulting from dark counts of the PMT.
The second switch enables selection between manual or gate-driven integration
of the acquisition window. The third switch activates low-pass filtering with a
cutoff frequency of \SI{40}{MHz}.

\subsection{Acquisition system}
The DAQ required redesigning for compatibility with the new VFC-HD carrier,
necessitating the integration of a Rear Transition Module (RTM) to facilitate
control of the gate and the built-in test LED. Figure~\ref{fig:bsra_daq}
illustrates the principle of DAQ functionality.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_daq.pdf}
    \caption{Ackquisition system schematics}
    \label{fig:bsra_daq}
\end{figure}
The new DAQ receives asynchronous data samples from a 500MSPS 14-bit ADC
(depicted as a violet rectangle), organized in blocks of four frames
every~\SI{8}{\nano s}, corresponding to one channel. Each frame contains 4 x
16-bit values. To compute integrals over~\SI{100}{\nano\second}, the system
employs data tagging. The tagging block resamples LHC timing at 500MSPS and
generates bunch and turn clock vector masks, storing information about both
clock phases. The 14-bit ADC data is left-aligned to 16-bit values. The tagger
sets the two least significant bits to one when the ADC sample falls within the
edge range.\\
The system stores data from the region of interest into a FIFO using the
synchronization block pulse. The stored data is then processed as a stream of
values passed to the integrator. The integrator begins summing values upon
encountering the tag indicating the start of the bunch, generating 64 values
offset from the pedestal memory (P[63:0]). This process subtracts accumulated
charge at the gate opening and closing edges. The resulting values are then
sent to the EMA, and the results are stored in the
block RAM accessible via the VME64x bus. The format of the data provided to the
PC via the running FESA server matches that of the current
AGM~\cite{my_thesis}.

\subsection{Installation in LHC}

The instrument OF and DAQ are situated in LHC Point 4 near the RF system. The
new installation is depicted in Fig.~\ref{fig:installation}, delineated by a
violet frame. Adjacent to its right side is the currently utilized AGM, while
the BSRS occupies the left side. All instruments share the same light source
with the BSRT. The instrument is extracted from the shielding, with the red
arrow indicating the direction from which light emanates and enters the MCP-PMT
of the new AGM.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{installation.jpg}
    \caption{New AGM installed in Point 4 next to BSRT}
    \label{fig:installation}
\end{figure}
Atop the reflective foil, which directs light towards the MCP-PMT, there's a
LED serving to simulate the light source, aiding in system verification. The
AGM amplifier is positioned atop the optical front-end for easy access during
interventions. To regulate light intensity, pneumatically controlled optical
filters are positioned in front of the MCP-PMT.

\section{Control Interface}
The gateware outputs data from four distinct block RAMs. The first RAM stores
raw data from the ADC, crucial for accurately setting the initial timing delay
of the instrument. It holds data for 1.5 turns, resulting in a publication rate
of~\SI{3.3}{Hz}. The subsequent two memories store both snapshot and averaged
data with finer granularity. They retain 10 samples per~\SI{10}{ms} window,
yielding a sample rate of~\SI{100}{Hz}. The final memory exclusively stores
averaged data collected every~\SI{10}{ms}, dictating the publication rate. Data
retrieval is managed by the FESA server, which receives interrupts whenever the
data becomes available.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{control_interface.jpg}
    \caption{New AGM expert graphical interface}
    \label{fig:control_interface}
\end{figure}
To streamline setup tuning, a new interface was developed. Operating as a
client of the FESA server, it showcases the content of all block RAMs. The
primary control panel is illustrated in Fig.~\ref{fig:control_interface}. It
features three plots: raw data from the ADC (top plot), processed averaged data
(bottom-left plot), and the temporal evolution of the data (bottom-right plot).
The left menu governs amplifier settings, while the bottom menu regulates
acquisition delays and optical filters.

\section{First acquired plots}
The initial measurements were conducted in November 2023 using the ion beam.
The optical system divides the SL in a 1:8 ratio, resulting in significantly
less light reaching the new system compared to the current one. Consequently,
we anticipate lower deviation values for the system measuring light with higher
intensity.
\begin{figure}[!htb]
    \begin{center}
        \scalebox{0.54}{\input{plots/ions.pgf}}
        \caption{Features comparison of AGMs}
        \label{fig:comparison_chart}
    \end{center}
\end{figure}
Figure~\ref{fig:comparison_chart} displays the initial acquisition results of
the new system compared to the current one. The red plot represents
measurements by the current system, while the blue plot depicts the new
acquisition. Within the range of the 47th to 63rd bin, the gate is closed,
resulting in white noise. However, the red plot gradually decays and extends
below the zero line due to artifacts stemming from analog integrators,
exhibiting non-linear behavior. The new system effectively eliminates this
artifact. The yellow plot illustrates the value deviation of the new
acquisition from the current system, with the relative value deviation depicted
as well. In this measurement, the maximum deviation is~4\% in the region
corresponding to the decay of the red plot. Additionally, a deviation of~2\% in
the steep sections may be attributed to sample phase misalignment between the
two systems.

\section{Summary}
The new AGM has been integrated onto the CERN BI standard platform VFC-HD
carrier, streamlining maintenance procedures. By eliminating IBMS cards
featuring the lhcb2002b ASIC, the design has addressed issues related to
cumbersome calibration and artifacts in the output plots. The changes in the
AGM design compared to the current AGM are summarized in
Table~\ref{tab:system_comparison}.\\
The system is installed at LHC Point 4, alongside the current AGM, collecting
data for further analysis. To conduct a comprehensive reliability test and
verify proper functionality, additional data collection is required. Initial
observations indicate that despite receiving less light, the new system can
still achieve comparable performance to the current AGM.
\input{tables/system_comparison.tex}
